CommonMind Consortium: DLPFC

### 2020-10-12 combined_analysis/CTS/GABA.Rmd
initial commit

---
title: "CommonMind Consortium: DLPFC"
subtitle: "Cell type specific analysis of read count data by cohort"
author:
- name: "[Gabriel Hoffman](http://gabrielhoffman.github.io)$^{1}$ and Kelsey Montgomery$^{2}$"
  affiliation: | 
    [1] Icahn School of Medicine at Mount Sinai, New York
    [2] Sage Bionetworks, Seattle, WA
date: "Run on `r Sys.time()`"
documentclass: article
output:
  html_document:
    toc: true
    toc_float: true
params:
  upload: FALSE
---

@@

rmarkdown::render("combined_analysis/combine_all_DLPFC_by_celltype.Rmd")

@@

```{r knit2synapse, eval= FALSE}
library(synapser)
library(knit2synapse)
synLogin()
knit2synapse::createAndKnitToFolderEntity(file = "combined_analysis/combine_all_DLPFC_by_celltype.Rmd",
                                          parentId ="syn22416298",
                                          folderName = 'Cell type specific analysis of read count data by cohort',
                                          overwrite=TRUE, 
                                          knitmd=TRUE)
```

@@

library(githubr)
parentId = 'syn22416298';
activityName = 'Cell type specific analysis of read count data by cohort';
activityDescription = 'Cell type specific analysis of read count data by cohort';

thisFileName <- 'combine_all_DLPFC_by_celltype.Rmd'

# Github link
thisRepo <- getRepo(repository = "CommonMindConsortium/covarr-de", ref="branch", refName='master')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('combined_analysis/',thisFileName))
```

@@

cellType = "GABA"
geneExpr = geneExpr.CTS[,cellType,]
geneExpr.se = geneExpr.CTS.se[,cellType,]

@@

## Summarize samples
```{r summary}
METADATA %>% group_by(Dx, Institution) %>% dplyr::summarise(count = n()) %>% 
  kable(caption="Diagnosis, Institution") %>% kable_styling(full_width=FALSE)

METADATA %>% group_by( Institution, Reported_Gender) %>% dplyr::summarise(count = n()) %>% 
  kable(caption="Institution, Sex") %>% kable_styling(full_width=FALSE)

METADATA %>% group_by(Dx, Institution, Reported_Gender) %>% dplyr::summarise(count = n()) %>% 
kable(caption="Diagnosis, Institution, Sex") %>% kable_styling(full_width=FALSE) 
```

```{r joint.cca, fig.width=7, fig.height=7}
variables = c('Reported_Gender', 'CellType', 'ageOfDeath', 'Institution', "Dx")

df_cca = lapply(variables, function(v1){

  if( v1 == "CellType"){
    form1 = ~ GABA + GLU + Olig + MgAs
  }else{
    form1 = as.formula(paste0('~', v1))
  }
  M1 = model.matrix(form1, METADATA)

  res = lapply(variables, function(v2){

    if( v2 == "CellType"){
      form2 = ~ GABA + GLU + Olig + MgAs
    }else{
      form2 = as.formula(paste0('~', v2))
    }
    M2 = model.matrix(form2, METADATA)

    fit = cancor(M1, M2)
    V_sqrt = sqrt(mean(fit$cor^2)) # Cramer's sqrt V

    data.frame(v1, v2, V_sqrt)
  })
  do.call(rbind, res)
})
df_cca = do.call(rbind, df_cca)

C = matrix(df_cca$V_sqrt, ncol=length(variables))
rownames(C) = variables
colnames(C) = variables

plotCorrMatrix(C, main='CCA')
```

```{r explain.CellType}
CTs = c('GABA', 'GLU', 'Olig', 'MgAs')

form = ~ ageOfDeath:Institution + Reported_Gender + Dx + Institution

Y = clr(METADATA[,CTs])

df = lapply( colnames(Y), function(id){

  fit = lm(Y[,id] ~ ageOfDeath:Institution + Reported_Gender + Dx + Institution, METADATA)

  data.frame( Test     = rownames(anova(fit))[1:4],
              CellType = id, 
              p        = anova(fit)[,'Pr(>F)'][1:4], 
              frac     =  calcVarPart(fit)[1:4])
})
df = data.table(do.call(rbind, df))

df$Test[df$Test=='Reported_Gender'] = "Sex"
df$Test[df$Test=='ageOfDeath:Institution'] = "Age of death"
df$p.adj = p.adjust(df$p, "bonferroni")

ymax = max(df[,sum(frac),by="CellType"]$V1)*105
ggplot(df, aes(CellType, frac*100, fill=Test)) + geom_bar(stat="identity") + coord_flip() + theme_bw() + theme(aspect.ratio=1, panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) +  ylab("Variance explained (%)") + theme(axis.line = element_line(colour = "transparent"), 
        axis.line.x = element_line(colour = "black"), panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), panel.border = element_blank(), 
        panel.background = element_blank(), axis.ticks.y = element_blank(), 
        legend.key = element_blank(), plot.margin = unit(c(0, 
            0.3, 0, 0.8), "cm")) + scale_y_continuous(expand=c(0,0), limits=c(0, ymax)) + xlab('')

ggplot(df, aes(CellType, Test, fill=-log10(pmax(1e-16,p)), label=ifelse(p.adj<0.05, '*', ''))) + geom_tile() + theme_bw() + scale_fill_gradient(name = bquote(-log[10]~P), low="white", high="red", limit=c(0, NA)) + theme(aspect.ratio=1, panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) + geom_text() + coord_flip()
```

```{r sex.graph, fig.width=7, fig.height=7, dev="png"}
df_cca2 = df_cca[with(df_cca, v1!=v2),]
df_cca2$key = apply(df_cca2, 1, function(x){
  x = sort(c(x['v1'], x['v2']))
  paste(x[1], x[2])
  }) 
df_cca2 = df_cca2[!duplicated(df_cca2$key),]

rstat_nodes <- data.frame(name = unique(c(df_cca2$v1, df_cca2$v2)))

rstat_edges <- data.frame(from = match(df_cca2$v1, rstat_nodes$name),
                         to = match(df_cca2$v2, rstat_nodes$name),
                         Rsq = df_cca2$V_sqrt^2 )

net = tbl_graph(nodes = rstat_nodes, edges = rstat_edges, directed=FALSE)

ggraph(net,  layout = 'igraph', algorithm = 'star') + 
  geom_edge_link(aes(width = Rsq, color=Rsq), alpha = 1) +
  geom_node_point(size=20, color="grey70") +
  geom_node_text(aes(label = name), repel = FALSE) +
  theme_graph() + theme(aspect.ratio=1, plot.title = element_text(hjust = 0.5)) +
  scale_edge_width(name=bquote(R^2)) +
  scale_edge_color_gradient(name=bquote(R^2), low="white", high="red", lim=c(0, max(df_cca2$V_sqrt^2))) +  ggtitle("Pairwise squared correlation")
```

```{r sex.confounding, fig.width=5, fig.height=5}
form = ~ Reported_Gender + GABA + GLU + Olig + MgAs + ageOfDeath 
par(mfrow=c(2,2))
res = lapply( unique(METADATA$Institution), function(INST){
  i = METADATA$Institution == INST
  C = canCorPairs(form, METADATA[i,]) 
  plotCorrMatrix(C, main=INST)
})
```

```{r sex.by.age}
ggplot(METADATA, aes(Reported_Gender, ageOfDeath, fill=Reported_Gender)) + geom_boxplot() + theme_bw() + theme(aspect.ratio=1, plot.title = element_text(hjust = 0.5), legend.position="none") + facet_wrap(~Institution, nrow=1) + scale_fill_manual(values = c("red", "blue"))
```

```{r fraction.female}
df = xtabs(~Reported_Gender + Institution, METADATA)
df2 = apply(df, 2, function(x) x / sum(x))

plotPercentBars(data.frame(t(df2))) + scale_fill_manual(values=c("red", "blue")) + ylab("Fraction of samples") + xlab("Institution")
```

### Sex by cell fraction
```{r summary.sex, fig.width=8, fig.height=10}
figList = lapply(colnames(df_cellFractions), function(CT){
  ggplot(METADATA, aes_string('Reported_Gender', CT, fill='Reported_Gender')) + geom_boxplot() + theme_bw() + theme(aspect.ratio=1, plot.title = element_text(hjust = 0.5)) + facet_wrap(~Institution, nrow=1) + scale_fill_manual(values = c("red", "blue"))
})
plot_grid(plotlist=figList, ncol=1) 
```

### Age by cell fraction
```{r summary.Age, fig.width=8, fig.height=10}
figList = lapply(colnames(df_cellFractions), function(CT){
  ggplot(METADATA, aes_string('ageOfDeath', CT)) + geom_point() + geom_smooth(method="gam") + theme_bw() + theme(aspect.ratio=1, plot.title = element_text(hjust = 0.5)) + facet_wrap(~Institution, nrow=1)
}) 
plot_grid(plotlist=figList, ncol=1) 
```

## Age by cell fraction log ratio
```{r summary.Age.log.ratio, fig.width=8, fig.height=8}
figList = lapply(colnames(celFrac_ilr), function(CT){
  ggplot(METADATA, aes_string('ageOfDeath', CT)) + geom_point() + geom_smooth(method="gam") + theme_bw() + theme(aspect.ratio=1, plot.title = element_text(hjust = 0.5)) + facet_wrap(~Institution, nrow=1)
})
plot_grid(plotlist=figList, ncol=1) 
```

@@

```{r store, eval=TRUE}
if( params$upload ){    
  # Code
  # Make sure these match the folder name and parent Id in the first chunk.
  CODE <- Folder(name = "Cell type specific analysis of read count data by cohort", parentId = "syn22416298")
  CODE <- synStore(CODE)

  # Store data 
  storeData = function( name, description){
    file = paste0(name, ".RDS")
    saveRDS( eval(parse(text=name)), file=file)

    nEXP_OBJ = File(file, 
                    name = description, 
                    parentId = CODE$properties$id)
    synStore(nEXP_OBJ, used = ALL_USED_IDs, executed = thisFile)
  }

>>>>

---
title: "CommonMind Consortium: DLPFC"
subtitle: "Cell type specific analysis of read count data by cohort"
author:
- name: "[Gabriel Hoffman](http://gabrielhoffman.github.io)$^{1}$ and Kelsey Montgomery$^{2}$"
  affiliation: | 
    [1] Icahn School of Medicine at Mount Sinai, New York
    [2] Sage Bionetworks, Seattle, WA
date: "Run on `r Sys.time()`"
documentclass: article
output:
  html_document:
    toc: true
    toc_float: true
params:
  upload: FALSE
  cellType: "GABA"
---

@@

rmarkdown::render("combined_analysis/CTS/GABA.Rmd")

@@

```{r knit2synapse, eval= FALSE}
library(synapser)
library(knit2synapse)
synLogin()
knit2synapse::createAndKnitToFolderEntity(file = "combined_analysis/combine_all_DLPFC_by_celltype.Rmd",
                                          parentId ="syn23241817",
                                          folderName = 'Analysis of read count data by cohort: GABA',
                                          overwrite=TRUE, 
                                          knitmd=TRUE)
```

@@

library(githubr)
parentId = 'syn23241817';
activityName = 'Analysis of read count data by cohort: GABA';
activityDescription = 'Analysis of read count data by cohort: GABA';

thisFileName <- 'GABA'

# Github link
thisRepo <- getRepo(repository = "CommonMindConsortium/covarr-de", ref="branch", refName='master')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('combined_analysis/CTS/',thisFileName))
```

@@

geneExpr = geneExpr.CTS[,params$cellType,]
geneExpr.se = geneExpr.CTS.se[,params$cellType,]

@@

@@

```{r store, eval=TRUE}
if( params$upload ){    
  # Code
  # Make sure these match the folder name and parent Id in the first chunk.
  CODE <- Folder(name = activityName, parentId = parentId)
  CODE <- synStore(CODE)

  # Store data 
  storeData = function( name, description){
    file = paste0(name, "_", params$cellType, ".RDS")
    saveRDS( eval(parse(text=name)), file=file)

    nEXP_OBJ = File(file, 
                    name = paste0(params$cellType, ' - ',description), 
                    parentId = CODE$properties$id)
    synStore(nEXP_OBJ, used = ALL_USED_IDs, executed = thisFile)
  }

### 2020-10-12 combined_analysis/CTS/MgAs.Rmd
initial commit

  # Store data 
  storeData = function( name, description){
    file = paste0(name, ".RDS")
    saveRDS( eval(parse(text=name)), file=file)

    nEXP_OBJ = File(file, 
                    name = description, 
                    parentId = CODE$properties$id)
    synStore(nEXP_OBJ, used = ALL_USED_IDs, executed = thisFile)
  }

>>>>

  # Store data 
  storeData = function( name, cellType, description){
    file = paste0(name, "_", cellType, ".RDS")
    saveRDS( eval(parse(text=name)), file=file)

    nEXP_OBJ = File(file, 
                    name = paste0(cellType, ' - ',description), 
                    parentId = CODE$properties$id)
    synStore(nEXP_OBJ, used = ALL_USED_IDs, executed = thisFile)
  }
